Gaussian Process regression for astronomical time-series

### 2022-11-08 src/tex/ms.tex
[showyourwork] overleaf sync

Gaussian Processes (GPs) are a powerful class of statistical models, which allow us to define a probability distribution over random functions. Rather than write down an explicit mathematical formula for the function from which some observations are generated, we model the covariance between pairs of samples from the process, using our physical domain knowledge and/or available data to guide our modelling.
While this may seem abstract, GPs have a wide range of applications from modelling of stochastic physical processes, to high dimensional interpolation and smoothing.
\begin{armarginnote}[]
  \entry{GP}{Gaussian Process}
  \entry{GPR}{Gaussian Process Regression}
\end{armarginnote}
In particular, GP Regression (GPR) has become increasingly popular in the astronomical community over the last decade. Part of the reason for this uptake is the growing availability and importance of time-domain datasets in astronomy. These systematically contain non-trivial random or unknown signals, whether astrophysical or instrumental, that need to be modelled. In many cases, these are nuisance signals, which we need to marginalise over in order to detect or measure other signals robustly. Sometimes, we are interested in the stochastic behaviour itself, and want to infer its characteristics or predict its behaviour. GPR offers a compelling solution: statistically principled, naturally Bayesian, and extremely flexible, yet mathematically simple. To new users, however, the lack of an explicit functional form for the model can make GPR can seem a little arcane. Furthermore, the computational cost of the method, which scales cubically with the dataset size, can be an obstacle. These factors initially impeded its dissemination in the astronomical community, but have been largely overcome in recent years thanks to the availability of user-friendly, computationally optimised software packages.

@@

A number of important factors have contributed to recent the democratisation of GPR across a wide range of scientific disciplines, including the publication of a dedicated textbook \citep{gpml}, as well as the availability of cheap computing power and user-friendly, open-source GPR software. In astronomy, specifically, an additional factor has been at play: the rise of time domain surveys. Correlated noise in time-domain observations is a direct and unavoidable consequence of causation, and hence ubiquitous. Adequately modelling this correlated noise is vital when searching for faint signals, for example from exoplanets. Astrophysical sources, from accretion disks on all scales to magnetically active stars or cloudy brown dwarfs, also display complex, intrinsically or apparently stochastic behaviour, for which adequate modelling strategies are required. GPR is a natural choice to tackle these challenges. Rather than attempting to cover all applications of GPs in astrophysics, which would not be feasible in the space available, we have therefore opted to focus on its application to time-domain datasets in astronomy.

@@

@@

In this second example, we demonstrate another common application of GPR in astrophysics, as a flexible model for nuisances and correlated noise where we want to correctly capture uncertainty introduced by this noise model into our constraints on the parameters of interest.
To this end, we simulate the light curve of a transiting exoplanet and aim to infer the physical parameters of the system, taking correlated noise into account.
The correlated noise in transit light curves is typically caused by the variability of the host star and by instrumental effects like focus or pointing changes.
It has been demonstrated that neglecting to account for these variations can cause significant errors when inferring the properties of the planet \citep{2006MNRAS.373..231P,2007A&A...472L..13G}.
This correlated noise can often be well modelled by a GP, and the use of GPs for transit modelling has been a fruitful area of research in the astrophysics literature (see Section~\ref{sec:transit_fit} for a more detailed discussion).

@@

We then used the same framework to model the simulated dataset, first ignoring the correlated noise and then accounting for it explicitly.
For the transit model, we fit for the planet-to-star radius ratio $R_{\rm p}/R_\star$, the mid-transit time $T_0$, the out-of-transit flux $f_0$, and the limb darkening parameters $u_1,\,u_2$.
When also fitting for correlated noise, we also marginalise over the amplitude $\amplitude$ and timescale $\lengthscale$ of the GP model.
The posterior constraints on the time of transit and the planet-to-star ratio are shown in the right panel of \autoref{fig:transit}, with the true values of these parameters over-plotted as a black line.
When the correlated noise is neglected (plotted in orange in \autoref{fig:transit}), the inferred parameters are significantly inconsistent with the truth.
Taking the correlated noise into account (blue in \autoref{fig:transit}) increases the uncertainties on the inferred parameters, but also shifts the results to recover the true parameters.

@@

The remainder of this review is structured as follows. Section~\ref{sec:basics} provides a brief, accessible introduction to the basic theory of GPR, and Section~\ref{sec:choices} discusses the key modelling choices one needs to make when using a GP to model data, with practical advice for how to make these choices. Section~\ref{sec:uses} gives an overview of applications of GPR to time-domain astronomical datasets to date, from stars and exoplanets to active galactic nuclei, pulsars and gravitational waves. In Section~\ref{sec:challenges} we discuss some challenges of using GPR in practice, and how they may be overcome, while Section~\ref{sec:open} reviews the ecosystem of open-source GPR software available today. Finally, we summarise the main take-away points and outline future directions in Section~\ref{sec:concl}.

@@

\begin{figure}[ht]
  \centering
  \script{GPR_PGM.ipynb}
  \includegraphics[width=0.8\linewidth]{figures/GPR_PGM.pdf}
  \caption{Probabilistic Graphical Model (chain graph) for GP regression (adapted from Figure 2.3 of \citealt{gpml}) given observations $\mathbf{y}$ at inputs $\mathbf{t}$, and a prediction at test input $t_\star$. Dots represent fixed variables (inputs), grey circles represent observed variables white circles represent unknown (latent) variables. The thick horizontal bar indicates a set of fully connected nodes (meaning that every $f$ depends on every $t$). On the other hand, each observed output $y_i$ depends only on the corresponding $f_i$ and $\sigma_i$, and is conditionally independent of the other variables. In this example, the measurement uncertainties $\boldsymbol{\sigma}$ and the GP hyper-parameters (not shown) are treated as known (fixed).}
  \label{fig:GPR_PGM}
\end{figure}

The parameters $\meanparams$ and $\hyperparams$ of the mean and covariance function are known as the \textit{hyper-parameters} of the GP. Strictly speaking, the \textit{parameters} of the GP are the (infinitely many) unknown functions that share the specified mean vector and covariance matrix and could have given rise to the observations. However, these parameters are always marginalised over: we never explicitly deal with the individual functions, except when drawing samples for illustrative purposes (as we did in Figures~\ref{fig:quasar} and \ref{fig:transit}). GPs are therefore a type of \textit{Hierarchical Bayesian Model} (HBM). To illustrate this, we show a Probabilistic Graphical Model (PGM) of GPR in Figure~\ref{fig:GPR_PGM}. In this type of graph, shaded circles represent random variables that are observed, while un-shaded circles represent latent, or unobserved, random variables. We never observe the unknown function that gave rise to the data directly, but infer a probability distribution for it from our noisy observations.

@@

The formal definition given in Equation~\ref{eq:gp_def} does not necessarily provide an intuitive understanding of how GPs work. Most readers of this review will, however, be more familiar with least-squares regression. In this section we will show that GPR can be thought of as a generalisation of least-squares regression, allowing for correlated noise (or signals) in the data. Conversely, least-squares regression, as traditionally presented, is a special case of GPR, where the covariance matrix is assumed to be purely diagonal, and the variances associated with each observation are known \textit{a priori}.

@@

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function}, or \emph{kernel function}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~\ref{eq:gp_like}. Depending on the choice of kernel function and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

The kernel function $k$ encodes our beliefs about the stochastic, or random, element of the model, in just the same way as the mean function $m$ encodes our beliefs about the deterministic component of the model. For example, in many circumstances, we would expect that two observations taken close together in time should be more strongly correlated than observations taken further apart, and we would use a decreasing function of the time interval $\dt=|t_i-t_j|$ to represent that. One of the most widely used covariance functions is the squared exponential
\begin{equation}
\label{eq:kse}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,\lengthscale^2}\right).
\end{equation}
This gives rise to smooth random functions with variance $\amplitude$ and characteristic length-scale $\lambda$. We present other commonly used kernel functions and discuss how to select one in Section~\ref{sec:kernels}.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{figures/GPR_workflow.pdf}
  \caption{Schematic representation of a typical GPR workflow. Given a dataset (blue box) and some modelling choices (grey boxes, Section~\ref{sec:choices}), the mathematical framework presented in Section~\ref{sec:basics} can be used (green boxes) to evaluate the likelihood and posterior distribution over the hyper-parameters which can be optimised or sampled. We can also condition the GP on the data to predict observations at locations where we do not yet have observations. Several of the plots in this figure are taken from the 
  tutorial on hyper-parameter optimization included in the {\sc george} package documentation (\url{https://george.readthedocs.io}), and show the composite GP model proposed by \citet{gpml} for the Mauna Loa CO$_2$ dataset of \citet{MaunaLoaCO2}.}  
  \label{fig:GPR_workflow}
\end{figure}

Now that we know how to evaluate the likelihood (Equation~\ref{eq:gp_like}), we are ready to perform \textit{inference}, that is to use observations to update our prior beliefs about the system we are observing. The overall Bayesian inference workflow for GPR regression is illustrated schematically in Figure~\ref{fig:GPR_workflow}. Given a dataset (blue, left panel), our first task is typically to select a kernel function, along with an educated initial guess for its hyper-parameters ($2^{\rm nd}$ column, top panel). These choices are discussed in more detail in Section~\ref{sec:choices}, and should be guided by a combination of domain knowledge (what we know about the process that generated the data), practical considerations (e.g.\ ease of implementation and computational cost) and diagnostics based on the data themselves (e.g.\ examining the auto-correlation function and/or the power spectral density of the data). Given a kernel function and hyper-parameters, we can draw samples from the GP prior ($3^{\rm rd}$ column, top panel), which can serve as sanity check for the choice of kernel function (in the sense that, measurement errors aside, samples from the GP prior should ideally look qualitatively similar to the observations).

We are now in a position to compute the GP covariance matrix for the data ($2^{\rm nd}$ column, $2^{\rm nd}$ panel), and to evaluate the likelihood of the model ($2^{\rm nd}$ column, $3^{\rm rd}$ panel), which we can then optimise with respect to the hyper-parameters ($3^{\rm rd}$ column, $2^{\rm nd}$ panel) using standard optimization methods; this is called \emph{training} the GP. Often, however, we would incorporate priors on the hyper-parameters ($2^{\rm nd}$ column, bottom panel), which we select once again based on domain knowledge and practical considerations. We can then evaluate the \emph{posterior} distribution ($3^{\rm rd}$ column, $3^{\rm rd}$ panel):
\begin{equation}
  p(\meanparams,\hyperparams|\mathbf{y})=\frac{p(\mathbf{y}|\meanparams,\hyperparams)p(\meanparams,\hyperparams)}{p(\mathbf{y})},
\end{equation}
where the denominator $p(\mathbf{y})=\int p(\mathbf{y}|\meanparams,\hyperparams)p(\meanparams,\hyperparams) \mathrm{d}\meanparams \mathrm{d}\hyperparams$ is the marginal likelihood, or model evidence, and is often omitted from calculations since it is constant, but becomes relevant for model comparison. Once again, we can optimise this posterior with respect to the hyper-parameters ($3^{\rm rd}$ column, $2^{\rm nd}$ panel). For a given set of hyper-parameters (for example, the set which maximises either the likelihood or the posterior), the GP prior can be conditioned on the observations to make predictions ($4^{\rm th}$ column, top panel); this is called \emph{conditioning} the GP, and is discussed in Section~\ref{sec:pred}.

However, in astronomical applications we often need to estimate the full posterior distribution over the hyper-parameters. 
Generally we cannot do this analytically, so we resort to sampling methods such as Markov Chain Monte Carlo (MCMC) or nested sampling ($3^{\rm rd}$ column, bottom panel). The samples can be used to estimate posterior distributions for individual hyper-parameters of interest ($4^{\rm th}$ column, $2^{\rm nd}$ panel). For example, we might be using a GP to take into account correlated noise when fitting an otherwise deterministic model to some data, but we are not interested in the noise \emph{per se}, only on the impact it has on the parameters of the mean function, $\meanparams$. We would then \emph{marginalise} (integrate over, or project) the posterior samples over the \emph{nuisance parameters} $\hyperparams$. Some samplers also enable us to compute the model evidence, which can be used to compare different kernel functions (see Section~\ref{sec:model_selection} for a discussion). Finally,  posterior samples can also be used in a Monte Carlo fashion to compute predictions that incorporate the uncertainties on the hyper-parameters ($4^{\rm th}$ column, bottom panel).

@@

An important use case for GPR is as a statistically principled interpolation method. We wish to ``learn'' an unknown function that gave rise to some data, in order to make predictions for some new set of inputs. This process is illustrated schematically in Figure~\ref{fig:GPR_PGM}, where such a test input and the corresponding function value are highlighted in red. Importantly, a GP model provides not a point estimate but a full probability distribution for the function at any desired location(s) in the input domain. This allows for robust uncertainty propagation (though there are some important caveats we will touch upon in Section~\ref{sec:caution}), and can also motivate strategies for active sampling (deciding when or where to make new observations).

@@

Given some existing observations $\mathbf{y}$, taken at times $\mathbf{t}$, how do we make predictions at some new set of times $\mathbf{t}_\star$? We are after $p(\mathbf{y}_\star|\mathbf{y})$, the \textbf{conditional} probability distribution for $\mathbf{y}_\star$ given $\mathbf{y}$. In GPR this is also known as the \emph{predictive distribution}, because it is often used to extrapolate a time-series dataset forwards. The `magic' of GPR is that the predictive distribution is also Gaussian, and its mean and covariance are given by simple analytic relations:
\begin{equation}
  \label{eq:pred}
  \mathbf{f}_\star = \mathbf{m}_\star + \mathbf{K}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, (\mathbf{y} - \mathbf{m}) ~\mathrm{~~~and~~~}
  \mathbf{C}_\star =   \mathbf{K}_{\star\star} - \mathbf{K}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, \mathbf{K}_\star,
\end{equation}
where we have introduced the notation $\mathbf{m}_\star \equiv m(\mathbf{t}_\star,\meanparams)$, $\mathbf{K}_\star \equiv k(\mathbf{t},\mathbf{t}_\star, \hyperparams)$ and
$\mathbf{K}_{\star\star} \equiv k(\mathbf{t}_\star,\mathbf{t}_\star,\hyperparams)$. Note that, as real observations are always noisy, $\mathbf{K}$ generally includes a white noise term ($\delta_{ij} \sigma^2_i$), but $\mathbf{K}_\star$ does not. Depending on whether one wishes the predictive variance (the diagonal elements of $\mathbf{C}_\star$)  to account for measurement uncertainties or not, one can include the white noise term in $\mathbf{K}_{\star\star}$ or leave it out.

\subsubsection{Properties of the predictive distribution}
A closer look at Equation~\ref{eq:pred} reveals some important properties.
First, a GP is a \textit{linear predictor}. The predictive mean for a specific time $t_\star$ can be written as a linear combination of the observations:
$f_\star = \mathbf{w}^{\mathrm{T}} \, \mathbf{y}$, where $\mathbf{w} = \mathbf{K}^{-1} \, \mathbf{k}_\star$ and $\mathbf{k}_\star=k(\mathbf{t},t_\star,\hyperparams)$. It can also be written as
a linear combination of covariance functions centred on the training points:
$f_\star = \mathbf{\alpha}^\mathrm{T} \,  \mathbf{k}_\star$ where $\mathbf{\alpha} = \mathbf{K}^{-1} \, \mathbf{y}$.
These linearity properties are very important to understand the behaviour of GPs. They shed light on the relationship between GPR and standard linear models with very large numbers of free parameters \citep[see e.g.][for a more detailed discussion]{2021PASP..133i3001H},
and lead to a number of powerful extensions that are beyond  scope of this review.

Second, the predictive covariance is independent of the data. $\mathbf{C}_\star$ depends only on the \emph{locations} $\mathbf{t}$ of the observations, not on their values $\mathbf{y}$. This has important consequences for observation planning: if we know the covariance function $k$ and its parameters, we can decide when to observations to optimise our predictions at a given (set of) time(s). However, in practice we rarely know the hyper-parameters \textit{a priori}, and the predictive posterior distribution marginalised over the hyper-parameters \textit{does} depend on the observations.

@@

\begin{textbox}[htb]
  \section{Kernel function}
  The ``kernel'' or ``covariance'' function, typeset here as $k(t_i,\,t_j;\,\hyperparams)$ is a parametric description of the covariance between two data points $t_i$ and $t_j$. This function takes a pair of inputs (here $t_i$ and $t_j$) as arguments, and is controlled by a set of parameters $\hyperparams$, typically referred to as ``hyper-parameters''.

@@

% \section{FIRST-LEVEL HEADING}
% This is dummy text.
% % Heading 2
% \subsection{Second-Level Heading}
% This is dummy text. This is dummy text. This is dummy text. This is dummy text.

>>>>

Gaussian Processes (GPs) are a powerful class of statistical models, which allow us to define a probability distribution over random functions. Rather than write down an explicit mathematical formula for the function from which some observations are generated, we model the covariance between pairs of samples from the process, using our physical domain knowledge and/or available data to guide our modelling.
While this may seem abstract, GPs have a wide range of applications from modelling of stochastic physical processes, to high dimensional interpolation and smoothing.
\begin{armarginnote}[]
  \entry{GP}{Gaussian Process}
  \entry{GPR}{Gaussian Process Regression}
\end{armarginnote}
In particular, GP Regression (GPR) has become increasingly popular in the astronomical community over the last decade. Part of the reason for this uptake is the growing availability and importance of time-domain datasets in astronomy. These systematically contain non-trivial random or unknown signals, whether astrophysical or instrumental, that need to be modelled. In many cases, these are nuisance signals, which we need to marginalise over in order to detect or measure other signals robustly. Sometimes, we are interested in the stochastic behaviour itself, and want to infer its characteristics or predict its behaviour. GPR offers a compelling solution: statistically principled, naturally Bayesian, and extremely flexible, yet mathematically simple. To new users, however, the lack of an explicit functional form for the model can make GPR can seem a little arcane. Furthermore, the computational cost of the method, which naively scales cubically with the dataset size, can be an obstacle. These factors initially impeded its dissemination in the astronomical community, but have been largely overcome in recent years thanks to the availability of user-friendly, computationally optimised software packages.

@@

A number of important factors have contributed to recent the democratisation of GPR across a wide range of scientific disciplines, including the publication of a dedicated textbook \citep{gpml}, as well as the availability of cheap computing power and user-friendly, open-source GPR software. In astronomy, specifically, packages such as {\sc george} \citep{george} and {\sc celerite} \citep{celerite}, which were developed by astronomers working with applied mathematicians have been instrumental in broadening the user base for GPR. An additional factor has also been at play in this application domain: the rise of time domain surveys. Correlated noise in time-domain observations is a direct and unavoidable consequence of causation, and hence ubiquitous. Adequately modelling this correlated noise is vital when searching for faint signals, for example from exoplanets. Astrophysical sources, from accretion disks on all scales to magnetically active stars or cloudy brown dwarfs, also display complex, intrinsically or apparently stochastic behaviour, for which adequate modelling strategies are required. GPR is a natural choice to tackle these challenges. Rather than attempting to cover all applications of GPs in astrophysics, which would not be feasible in the space available, we have therefore opted to focus on its application to time-domain datasets in astronomy.

@@

It is worth noting that the determination of the time-delay for 0957+561 attracted considerable controversy at the time \citep[see][and references therein]{1995ApJ...455L...5K} and that later observations \citep{1997ApJ...482...75K} favour a significantly shorter time-delay than that obtained by \citet{prh92a}. Whatever the statistical method used, 
the likelihood function for translation problems of this type can be highly multi-modal on large scales, making identification of the global optimum challenging. Furthermore, the assumption that one quasar image's light curve is merely a scaled and shifted version of the other is probably too simplistic. This is why we opted to use simulated data in this example, rather than work with the  dataset originally analysed by \citet{prh92a}.

@@

In this second example, we demonstrate another common application of GPR in astrophysics, as a flexible model for nuisances and correlated noise where we want to correctly capture uncertainty introduced by this noise model into our constraints on the parameters of interest.
To this end, we simulate the light curve of a transiting exoplanet and aim to infer the physical parameters of the system, taking correlated noise into account.
The correlated noise in transit light curves is typically caused by the variability of the host star and by instrumental effects like focus or pointing changes.
It has been demonstrated that neglecting to account for these variations can cause significant errors when inferring the properties of the planet \citep{2006MNRAS.373..231P,2007A&A...472L..13G}.
This correlated noise can often be well modelled by a GP, and the use of GPs for transit modelling has been a fruitful area of research in the astrophysics literature (see \autoref{sec:transit_fit} for a more detailed discussion).

@@

We then used the same framework to model the simulated dataset, first ignoring the correlated noise and then accounting for it explicitly.
For the transit model, we fit for the planet-to-star radius ratio $R_{\rm p}/R_\star$, the mid-transit time $T_0$, the out-of-transit flux $f_0$, and the limb darkening parameters $u_1,\,u_2$.
When also fitting for correlated noise, we also marginalise over the amplitude $\amplitude$ and timescale $\lengthscale$ of the GP model.
The posterior constraints on the time of transit and the planet-to-star ratio are shown in the right panel of \autoref{fig:transit}, with the true values of these parameters over-plotted as a black line.
When the correlated noise is neglected (plotted in orange in \autoref{fig:transit}), the inferred parameters are significantly inconsistent with the truth.
In this fit, while the correlated noise is neglected, we fit for an excess variance or ``jitter'' term, which is a method commonly used to account for model misspecification.
Taking the correlated noise into account (blue in \autoref{fig:transit}) increases the uncertainties on the inferred parameters, but also shifts the results to recover the true parameters.

% Do we want to say something about model checking here? The white noise version is clearly a bad fit!

@@

The remainder of this review is structured as follows. \autoref{sec:basics} provides a brief, accessible introduction to the basic theory of GPR, and \autoref{sec:choices} discusses the key modelling choices one needs to make when using a GP to model data, with practical advice for how to make these choices. \autoref{sec:uses} gives an overview of applications of GPR to time-domain astronomical datasets to date, from stars and exoplanets to active galactic nuclei, pulsars and gravitational waves. In \autoref{sec:challenges} we discuss some challenges of using GPR in practice, and how they may be overcome, while \autoref{sec:open} reviews the ecosystem of open-source GPR software available today. Finally, we summarise the main take-away points and outline future directions in \autoref{sec:concl}.

@@

% \begin{figure}[ht]
%   \centering
%   \script{GPR_PGM.ipynb}
%   \includegraphics[width=0.8\linewidth]{figures/GPR_PGM.pdf}
%   \caption{Probabilistic Graphical Model (chain graph) for GP regression (adapted from Figure 2.3 of \citealt{gpml}) given observations $\mathbf{y}$ at inputs $\mathbf{t}$, and a prediction at test input $t_\star$. Dots represent fixed variables (inputs), grey circles represent observed variables white circles represent unknown (latent) variables. The thick horizontal bar indicates a set of fully connected nodes (meaning that every $f$ depends on every $t$). On the other hand, each observed output $y_i$ depends only on the corresponding $f_i$ and $\sigma_i$, and is conditionally independent of the other variables. In this example, the measurement uncertainties $\boldsymbol{\sigma}$ and the GP hyperparameters (not shown) are treated as known (fixed).}
%   \label{fig:GPR_PGM}
% \end{figure}

The parameters $\meanparams$ and $\hyperparams$ of the mean and covariance function are known as the \textit{hyperparameters} of the GP. Strictly speaking, the \textit{parameters} of the GP are the (infinitely many) unknown functions that share the specified mean vector and covariance matrix and could have given rise to the observations. However, these parameters are always marginalised over: we never explicitly deal with the individual functions, except when drawing samples for illustrative purposes (as we did in Figures~\ref{fig:quasar} and \ref{fig:transit}). GPs are therefore a type of \textit{Hierarchical Bayesian Model} (HBM). 
%To illustrate this, we show a Probabilistic Graphical Model (PGM) of GPR in Figure~\ref{fig:GPR_PGM}. In this type of graph, shaded circles represent random variables that are observed, while un-shaded circles represent latent, or unobserved, random variables. 
We never observe the unknown function that gave rise to the data directly, but infer a probability distribution for it from our noisy observations.

@@

The formal definition given in \autoref{eq:gp_def} does not necessarily provide an intuitive understanding of how GPs work. Most readers of this review will, however, be more familiar with least-squares regression. In this section we will show that GPR can be thought of as a generalisation of least-squares regression, allowing for correlated noise (or signals) in the data. Conversely, least-squares regression, as traditionally presented, is a special case of GPR, where the covariance matrix is assumed to be purely diagonal, and the variances associated with each observation are known \textit{a priori}.

@@

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function}, or \emph{kernel function}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by \autoref{eq:gp_like}. Depending on the choice of kernel function and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

The kernel function $k$ encodes our beliefs about the stochastic, or random, element of the model, in just the same way as the mean function $m$ encodes our beliefs about the deterministic component of the model. For example, in many circumstances, we would expect that two observations taken close together in time should be more strongly correlated than observations taken further apart, and we would use a decreasing function of the time interval $\dt=|t_i-t_j|$ to represent that. One of the most widely used covariance functions is the squared exponential
\begin{equation}
\label{eq:kse}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,\lengthscale^2}\right).
\end{equation}
This gives rise to smooth (infinitely differentiable) random functions with variance $\amplitude$ and characteristic length-scale $\lambda$. We present other commonly used kernel functions and discuss how to select one in \autoref{sec:kernels}.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{figures/GPR_workflow.pdf}
  \caption{Schematic representation of a typical GPR workflow. Given a dataset (blue box) and some modelling choices (grey boxes, \autoref{sec:choices}), the mathematical framework presented in \autoref{sec:basics} can be used (green boxes) to evaluate the likelihood and posterior distribution over the hyperparameters which can be optimised or sampled. Note that, for simplicity, we have assumed a zero mean function in this figure. We can also condition the GP on the data to predict observations at locations where we do not yet have observations. Several of the plots in this figure are taken from the 
  tutorial on hyperparameter optimization included in the {\sc george} package documentation (\url{https://george.readthedocs.io}), and show the composite GP model proposed by \citet{gpml} for the Mauna Loa CO$_2$ dataset of \citet{MaunaLoaCO2}.}  
  \label{fig:GPR_workflow}
\end{figure}

Now that we know how to evaluate the likelihood (\autoref{eq:gp_like}), we are ready to perform \textit{inference}, that is to use observations to update our prior beliefs about the system we are observing. The overall Bayesian inference workflow for GPR regression is illustrated schematically in \autoref{fig:GPR_workflow}. Given a dataset (blue, left panel), our first task is typically to select a kernel function, along with an educated initial guess for its hyperparameters ($2^{\rm nd}$ column, top panel). These choices are discussed in more detail in \autoref{sec:choices}, and should be guided by a combination of domain knowledge (what we know about the process that generated the data), practical considerations (e.g.\ ease of implementation and computational cost) and diagnostics based on the data themselves (e.g.\ examining the auto-correlation function and/or the power spectral density of the data). Given a kernel function and hyperparameters, we can draw samples from the GP prior ($3^{\rm rd}$ column, top panel), which can serve as sanity check for the choice of kernel function (in the sense that, measurement errors aside, samples from the GP prior should ideally look qualitatively similar to the observations).

We are now in a position to compute the GP covariance matrix for the data ($2^{\rm nd}$ column, $2^{\rm nd}$ panel), and to evaluate the likelihood of the model ($2^{\rm nd}$ column, $3^{\rm rd}$ panel), which we can then optimise with respect to the hyperparameters ($3^{\rm rd}$ column, $2^{\rm nd}$ panel) using standard optimization methods; this is called \emph{training} the GP. Often, however, we would incorporate priors on the hyperparameters ($2^{\rm nd}$ column, bottom panel), which we select once again based on domain knowledge and practical considerations. We can then evaluate the \emph{posterior} distribution ($3^{\rm rd}$ column, $3^{\rm rd}$ panel):
\begin{equation}
  p(\meanparams,\hyperparams|\mathbf{y})=\frac{p(\mathbf{y}|\meanparams,\hyperparams)p(\meanparams,\hyperparams)}{p(\mathbf{y})},
\end{equation}
where the denominator $p(\mathbf{y})=\int p(\mathbf{y}|\meanparams,\hyperparams)p(\meanparams,\hyperparams) \mathrm{d}\meanparams \mathrm{d}\hyperparams$ is the marginal likelihood, or model evidence, and is often omitted from calculations since it is constant, but becomes relevant for model comparison. Once again, we can optimise this posterior with respect to the hyperparameters ($3^{\rm rd}$ column, $2^{\rm nd}$ panel). For a given set of hyperparameters (for example, the set which maximises either the likelihood or the posterior), the GP prior can be conditioned on the observations to make predictions ($4^{\rm th}$ column, top panel); this is called \emph{conditioning} the GP, and is discussed in \autoref{sec:pred}.

However, in astronomical applications we often need to estimate the full posterior distribution over the hyperparameters. 
Generally we cannot do this analytically, so we resort to sampling methods such as Markov Chain Monte Carlo (MCMC) or nested sampling ($3^{\rm rd}$ column, bottom panel). The samples can be used to estimate posterior distributions for individual hyperparameters of interest ($4^{\rm th}$ column, $2^{\rm nd}$ panel). For example, we might be using a GP to take into account correlated noise when fitting an otherwise deterministic model to some data, but we are not interested in the noise \emph{per se}, only on the impact it has on the parameters of the mean function, $\meanparams$. We would then \emph{marginalise} (integrate over, or project) the posterior samples over the \emph{nuisance parameters} $\hyperparams$. Some samplers also enable us to compute the model evidence, which can be used to compare different kernel functions (see \autoref{sec:model_selection} for a discussion). Finally,  posterior samples can also be used in a Monte Carlo fashion to compute predictions that incorporate the uncertainties on the hyperparameters ($4^{\rm th}$ column, bottom panel).

@@

An important use case for GPR is as a statistically principled interpolation method. We wish to ``learn'' an unknown function that gave rise to some data, in order to make predictions for some new set of inputs. 
%This process is illustrated schematically in Figure~\ref{fig:GPR_PGM}, where such a test input and the corresponding function value are highlighted in red.
Importantly, a GP model provides not a point estimate but a full probability distribution for the function at any desired location(s) in the input domain. This allows for robust uncertainty propagation (though there are some important caveats we will touch upon in \autoref{sec:caution}), and can also motivate strategies for active sampling (deciding when or where to make new observations).

@@

Given some existing observations $\mathbf{y}$, taken at times $\mathbf{t}$, how do we make predictions at some new set of times $\mathbf{t}_\star$? We are after $p(\mathbf{y}_\star|\mathbf{y})$, the \textbf{conditional} probability distribution for $\mathbf{y}_\star$ given $\mathbf{y}$. In GPR this is also known as the \emph{predictive distribution}, because it is often used to extrapolate a time-series dataset forwards. The `magic' of GPR is that the predictive distribution is also Gaussian, and its mean and covariance are given by simple analytic relations:
\begin{equation}
  \label{eq:pred}
  \mathbf{f}_\star = \mathbf{m}_\star + \mathbf{K}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, (\mathbf{y} - \mathbf{m}) ~\mathrm{~~~and~~~}
  \mathbf{C}_\star =   \mathbf{K}_{\star\star} - \mathbf{K}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, \mathbf{K}_\star,
\end{equation}
where we have introduced the vector $\mathbf{m}_\star \equiv m(\mathbf{t}_\star;\meanparams)$ and the matrices $\mathbf{K}_\star$ and $\mathbf{K}_{\star\star}$, with elements
\begin{equation}
  \label{eq:Kstar}
    [\mathbf{K}_\star]_{ij} = k(t_i,\mathbf{t}_{\star,j}; \hyperparams) ~~~~~ \text{and} ~~~~~
    [\mathbf{K}_{\star\star}]_{ij} = k(\mathbf{t}_{\star,i},\mathbf{t}_{\star,j};\hyperparams).
\end{equation}
Note that, as real observations are always noisy, $\mathbf{K}$ generally includes a white noise term ($\delta_{ij} \sigma^2_i$), but $\mathbf{K}_\star$ does not. The white noise term is usually omitted from (the diagonal elements of) $\mathbf{K}_{\star\star}$, in which case the predictive covariance $\mathbf{C}_\star$ accounts only for the uncertainties in the function inferred with the GP model. When making predictions that represent hypothetical additional observations taken with the same setup as the training set, however, the white noise term should be included in $\mathbf{K}_{\star\star}$. % Add ref to cross validationa nd eqn 27

\subsubsection{Properties of the predictive distribution}
A closer look at \autoref{eq:pred} reveals some important properties.
First, a GP is a \textit{linear predictor}\footnote{%
  There is some subtlety with respect to the non-linear mean function so, in this section, we set the mean to zero or, equivalently, consider the \emph{residuals} away from the mean model as the ``data''.}.
The predictive mean for a specific time $t_\star$ can be written as a linear combination of the observations:
$f_\star = \mathbf{w}^{\mathrm{T}} \, \mathbf{y}$, where $\mathbf{w} = \mathbf{K}^{-1} \, \mathbf{k}_\star$ and $\mathbf{k}_\star=k(\mathbf{t},t_\star,\hyperparams)$. It can also be written as
a linear combination of covariance functions centred on the training points:
$f_\star = \mathbf{\alpha}^\mathrm{T} \,  \mathbf{k}_\star$ where $\mathbf{\alpha} = \mathbf{K}^{-1} \, \mathbf{y}$.
These linearity properties are very important to understand the behaviour of GPs, as they shed light on the relationship between GPR and other types of models, including standard linear models with very large numbers of free parameters \citep[see e.g.][for a more detailed discussion]{2021PASP..133i3001H}.
%and lead to a number of powerful extensions that are beyond  scope of this review.

Second, the predictive covariance is independent of the data. $\mathbf{C}_\star$ depends only on the \emph{locations} $\mathbf{t}$ of the observations, not on their values $\mathbf{y}$. This has important consequences for observation planning: if we know the covariance function $k$ and its parameters, we can decide when to observations to optimise our predictions at a given (set of) time(s). However, in practice we rarely know the hyperparameters \textit{a priori}, and the predictive posterior distribution marginalised over the hyperparameters \textit{does} depend on the observations.

@@

\begin{textbox}[htb]
  \section{Kernel function}
  The ``kernel'' or ``covariance'' function, typeset here as $k(t_i,\,t_j;\,\hyperparams)$ is a parametric description of the covariance between two data points $t_i$ and $t_j$. This function takes a pair of inputs (here $t_i$ and $t_j$) as arguments, and is controlled by a set of parameters $\hyperparams$, typically referred to as ``hyperparameters''.

@@
